United Monoids

### 2022-01-11 paper/main.tex
Paper revision

% Fix broken footnotes
\makeatletter
\renewcommand\@makefntext[1]{\leftskip=0.7em\hskip-0.7em\@makefnmark\hspace{1mm}#1}
\makeatother
\renewcommand{\footnotesize}{\small}
\setlength{\footnotesep}{3mm}

@@

\lstset{
  backgroundcolor=\color{white},     % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}; should come as last argument
  % identifierstyle=\color{red},
  basicstyle=\normalsize\ttfamily,      % the size of the fonts that are used for the code
  breakatwhitespace=false,              % sets if automatic breaks should only happen at whitespace
  breaklines=false,                     % sets automatic line breaking
  captionpos=b,                         % sets the caption-position to bottom
  abovecaptionskip=-3 mm,
  commentstyle=\itshape\color{graygreen}, % comment style
  % escapeinside={(:}{:)},             % if you want to add LaTeX within your code
  escapechar={!},
  % extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  % firstnumber=1000,                % start line enumeration with line 1000
  % frame=tb,                        % adds a frame around the code
  % keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{darkblue},     % keyword style
  language=Haskell,                  % the language of the code
  morekeywords={ Set, Tree, Leaf, Node, Applicative, fmap, liftA2, bimap, foldMap
               , traverse, mappend, pure, Foldable, Traversable, zero, one
               , isZero, Semiring, Semigroup, NonEmpty, sconcat, TSet,
               , SimplicialSet, TSimplicialSet, Graph, TGraph, LGraph
               , Map, IsString, fromString },
  deletekeywords={instance, data, where, class, filter, type, insert, delete, union, map},      % if you want to delete keywords from the given language
  emph={data, class, instance, where, type},
  emphstyle=\color{darkpink},
  numbers=none,                      % where to put the line-numbers; possible values are (none, left, right)
  % numbersep=5pt,                   % how far the line-numbers are from the code
  % numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  % rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  % showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  % showstringspaces=false,          % underline spaces within strings only
  % showtabs=false,                  % show tabs within strings adding particular underscores
  % stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{grayred},     % string literal style
  % tabsize=2,                       % sets default tabsize to 2 spaces
  % title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
  xleftmargin=10pt,
  aboveskip=8pt,
  belowskip=4pt
}

@@

\newcommand{\add}{\text{\raisebox{-0.1mm}{\large $\oplus$}}}
\newcommand{\mul}{\text{\raisebox{-0.1mm}{\large $\otimes$}}}
\newcommand{\zero}{\raisebox{-0.2mm}{\textcircled{0}}\xspace}
\newcommand{\one}{\raisebox{-0.2mm}{\textcircled{1}}\xspace}
\newcommand{\two}{\raisebox{-0.2mm}{\textcircled{2}}\xspace}
\newcommand{\dia}{\,\text{\raisebox{-0.3mm}{\Large $\diamond$}}\,}
\newcommand{\arr}{\,\text{$\rightarrow$}\,}
\newcommand{\ldia}{\text{\raisebox{-0.3mm}{\Large $\diamond$}}}

@@

\title{United Monoids}
\subtitle{Finding Simplicial Sets and Labelled Algebraic Graphs on Trees}
\author{Andrey Mokhov}
\affiliation{Jane Street}

@@

\emph{Approach:} We follow a fairly standard approach of parameterising a data
structure with a semiring of edge labels. The newly developed data structure is
both more general and also simpler: the two ways of composing graphs used in the
original formulation can now be obtained from a single operation by fixing the
semiring parameter to zero and one, respectively.

\emph{Knowledge:} By analysing the resulting data structure and formulating laws
for interpreting expressions of the corresponding type, we discover an unusual
algebraic structure, which we call ``united monoids'', that is, a pair of
monoids whose unit elements coincide. We believe that it is worth studying
united monoids in their full generality, going beyond graphs which prompted
their discovery. To that end, we characterise united monoids with a minimal set
of axioms, prove a few basic theorems, and discuss several notable examples.

@@

Graphs and various graph-like combinatorial structures are ubiquitous in
programming. There are numerous approaches to representing graphs, from good old
adjacency matrices~\cite{cormen2009introduction} to categorical graph
algebras~\cite{2010_selinger_survey}. This paper introduces a new purely
functional graph representation. It is built on elementary foundations
(trees and semirings), and provides an expressive language for modelling
graph-like structures.

@@

\begin{lstlisting}
size :: Tree s a -> Int
size = fold (const 1) (const (+))
\end{lstlisting}

\noindent
Here we ignore the labels of leaves and internal nodes: every leaf is
interpreted as~$1$, and every internal node simply adds up the sizes of its
children. In the rest of the paper, we will study much more interesting ways to
fold trees. In particular, we will see how the very same data type can serve us
when working with sets, preorders, various flavours of graphs, as well as
simplicial sets. More specifically, the contributions of this paper are as
follows.

@@

\vspace{-1mm}
\begin{itemize}
  \item The lack of any connectivity can be expressed by~\zero; e.g, zero
        network bandwidth.
  \item Addition~$\add$ is a commutative\footnote{
            An operation $\bullet$ is \emph{commutative} if
            $a \bullet b = b \bullet a$.
        } operation for combining connections in \emph{parallel}; for example,
        adding bandwidth of independent network links between two locations.
  \item Multiplication~$\mul$ is for combining connections in \emph{sequence}.
        Continuing our example, $\mul$ corresponds to taking the \emph{minimum}
        bandwidth of network links on route between two locations.
        Multiplication distributes\footnote{
            \emph{Distributivity} of $\otimes$ over $\oplus$ means
            $a \otimes (b \oplus c) = (a \otimes b) \oplus (a \otimes c)$ and
            $(a \oplus b) \otimes c = (a \otimes c) \oplus (b \otimes c)$.
        } over addition and has \zero as a zero\footnote{
            An element $z$ is a \emph{zero} (or an \emph{annihilating element})
            of an operation $\bullet$ if $z \bullet a = a \bullet z = z$.
        }.
  \item Finally, \one is used to model either a unit of connectivity or infinite
        connectivity, depending on the specific semiring. In the bandwidth
        semiring, we have $\text{\one} = \infty$.
\end{itemize}

@@

class Semiring s where -- In this paper, !\h{s}! in !\h{Tree s a}! is always a !\h{Semiring}!
    zero !\hspace{2.9mm}!:: s !\hspace{14.9mm}!-- Unit of !\hadd!, zero of !\hmul!
    isZero :: s -> Bool !\hspace{2.5mm}!-- Testing for zero is often easier than implementing !\h{Eq s}!
    (!\hadd!) !\hspace{4.45mm}!:: s -> s -> s !\hspace{1.4mm}!-- Associative, commutative
    one !\hspace{3.65mm}!:: s !\hspace{14.9mm}!-- Unit of !\hmul!
    (!\hmul!) !\hspace{4.45mm}!:: s -> s -> s !\hspace{1.4mm}!-- Associative, distributes over !\hadd!

@@

\begin{lstlisting}[float,label=lst-tree,xleftmargin=0pt,belowskip=-2mm,caption={
    API for constructing and manipulating trees.
}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
leaf :: a -> Tree s a -- A tree with a single leaf
leaf = Leaf
!\vspace{-1mm}!
leaves :: Semiring s => NonEmpty a -> Tree s a -- A tree of !\n{``}!disconnected!\n{''}! leaves
leaves = sconcat . NonEmpty.map Leaf -- Via !\h{Semigroup}!'s !\h{sconcat}!
!\vspace{-1mm}!
node :: s -> Tree s a -> Tree s a -> Tree s a -- Combine two trees into an !\h{s}!-labelled node
node = Node
!\vspace{-1mm}!
size :: Tree s a -> Int -- The number of leaves in a tree
size = length -- Via !\h{Foldable}!
!\vspace{-1mm}!
leafSet :: Ord a => Tree s a -> Set a -- The set of leaves of a tree
leafSet = fold Set.singleton (const Set.union)
!\vspace{-1mm}!
hasLeaf :: Eq a => a -> Tree s a -> Bool -- Test if a tree contains a leaf !\h{a}!
hasLeaf = elem -- Via !\h{Foldable}!
!\vspace{-1mm}!
prune :: Tree s (Maybe a) -> Maybe (Tree s a) -- Prune all leaves labelled with !\h{Nothing}!
prune = sequence -- Via !\h{Traversable}!
!\vspace{-1mm}!
filter :: (a -> Bool) -> Tree s a -> Maybe (Tree s a) -- Prune all leaves !\h{a}! with !\h{p a = False}!
filter p = traverse (\a -> if p a then Just a else Nothing) -- Via !\h{Traversable}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

\noindent
It is sometimes necessary to be able to express the \emph{empty tree}. To do
that, we simply wrap the \hcode{Tree} data type into \hcode{Maybe}. This is
illustrated by the function \hcode{prune} (Listing~\ref{lst-tree}), which prunes
all leaves labelled with \hcode{Nothing} and returns a possibly empty tree as a
result. By combining the empty tree and the operation \dia, we can extend our
type of trees from a semigroup to a \emph{monoid}, i.e., a semigroup with the
unit element.

When tree interpretations respect the associativity of \dia, \zero-labelled
trees become isomorphic to non-empty lists. This is evidenced by the functions
\hcode{toList} (from the \hcode{Foldable} type class in
Listing~\ref{lst-tree-std}) and \hcode{leaves} (Listing~\ref{lst-tree}). In the
subsequent sections, we will study data structures obtained by instantiating
\hcode{Tree s a} with various semirings \hcode{s}, and by adding more
requirements (on top of associativity) to tree interpretations.

@@

\noindent
As hinted by the type synonym's name, we will interpret values \hcode{TSet a} as
(non-empty) sets; to get a standard \hcode{Set a} from a \hcode{TSet a}, we can
simply reuse \hcode{leafSet} from Listing~\ref{lst-tree}.
In fact, as demonstrated in Listing~\ref{lst-set}, a large part of the standard
set API can be be obtained by reusing more general \hcode{Tree}-manipulating
functions defined earlier in~\S\ref{sec-tree}.

\begin{lstlisting}[label=lst-set,xleftmargin=0pt,belowskip=7pt,caption={
    Implementing a part of the standard \code{Data.S}\code{et} API with
    \hcode{TSet}.
}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
type TSet a = Tree () a -- Sets are trees with !\h{()}!-labelled internal nodes
!\vspace{-1mm}!
instance Semiring () where -- The trivial semiring with !\h{zero = one = ()}!
    zero !\hspace{6.1mm}!= ()
    isZero () = True
    () !\add! () !\hspace{3.55mm}!= ()
    one !\hspace{7mm}!= ()
    () !\mul! () !\hspace{3.55mm}!= ()
!\vspace{-1mm}!
singleton :: a -> TSet a !\hspace{32.18mm}\vrule\hspace{8mm}!insert :: a -> TSet a -> TSet a
singleton = Tree.leaf    !\hspace{35.1mm}\vrule\hspace{8mm}!insert = (!\hdia!) . singleton -- Via !\h{Semigroup}!
!\vspace{-1mm}!
fromList :: [a] -> Maybe (TSet a) -- The empty set is represented by !\h{Nothing}!
fromList = fmap Tree.leaves . NonEmpty.nonEmpty
!\vspace{-1mm}!
delete :: Eq a => a -> TSet a -> Maybe (TSet a) -- The result can be empty
delete a = Tree.filter (/= a)
!\vspace{-1mm}!
member :: Eq a => a -> TSet a -> Bool !\hspace{9.1mm}\vrule\hspace{8mm}!union :: TSet a -> TSet a -> TSet a
member = Tree.hasLeaf                 !\hspace{30.33mm}\vrule\hspace{8mm}!union = (!\hdia!) -- Via !\h{Semigroup}!
!\vspace{-1mm}!
size :: Ord a => TSet a -> Int
size = Set.size . leafSet -- We can't use !\h{Tree.size}! because it uses a non-idempotent !\h{fold}!
!\vspace{-1mm}!
cartesianProduct :: TSet a -> TSet b -> TSet (a, b)
cartesianProduct = liftA2 (,) -- Via !\h{Applicative}!
!\vspace{-1mm}!
filter :: (a -> Bool) -> TSet a -> Maybe (TSet a) -- The result can be empty
filter = Tree.filter
!\vspace{-1mm}!
map :: (a -> b) -> TSet a -> TSet b
map = fmap -- Via !\h{Functor}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

\noindent
In addition to associativity, the set interpretation of trees puts two new
requirements on every call site \hcode{fold leaf node}\footnote{
    We should use \hcode{fold} with care since Haskell's type system isn't
    powerful enough to check such requirements. A dependently typed language,
    such as Agda~\cite{2007_norell_agda}, could help us here.
}:

@@

\section{Simplicial sets}\label{sec-simplicial-set}

@@

A \emph{simplicial set}~\cite{friedman_simplicial_sets} is a set of
\emph{simplices} of various dimensions along with their incidence relation.
A \emph{0-simplex} is a just a point, or a \emph{vertex}, and it will correspond
to a leaf in our trees. A \emph{1-simplex} is a pair of connected vertices, or
an \emph{edge}; a \emph{2-simplex} is a filled-in \emph{triangle}; a
\emph{3-simplex} is a solid \emph{tetrahedron}; and so on. To represent
simplices of dimension higher than $0$, we will \emph{connect} their vertices
using \one-labelled internal nodes. The figure below shows the simplicial sets
corresponding to the trees \code{"a"}, \code{"a" $\,\harr\,$ "b"},
\code{("a" $\,\harr\,$ "b") $\,\harr\,$ "c"} and
\code{("a" $\,\hdia\,$ "b") $\,\harr\,$ "c"}, where
\code{($\harr$)}~\code{=}~\hcode{node one} (see Listing~\ref{lst-simplicial-set}).

\vspace{5mm}
\hfill\hspace{-4mm}\includesvg[scale=0.28]{simplicial-set-examples}\hfill
\vspace{5mm}

\noindent
Note the difference between the last two examples: by changing the label of one
of the nodes from \one to \zero, we break the corresponding connection,
and as a result, a 2-simplex (triangle $\textit{abc}$) falls apart into
one-dimensional simplices (edges $\textit{ac}$~and~$\textit{bc}$).

@@

\begin{itemize}
    \item Associativity, inherited from~\S\ref{sec-tree}:
          $(a \dia b) \dia c = a \dia (b \dia c)$,
          $(a \arr b) \arr c = a \arr (b \arr c)$;
    \item Commutativity and idempotence, inherited from~\S\ref{sec-set}:
          $a \dia b = b \dia a$ and $a \dia a = a$;
    \item Containment (\textbf{\color{darkblue}new!}):
          $a \arr b = (a \arr b) \dia a \dia b$ (a simplex \emph{contains} its sub-simplices);
    \item Distributivity (\textbf{\color{darkblue}new!}):
          $a \arr (b \dia c) = (a \arr b) \dia (a \arr c)$,
          $(a \dia b) \arr c = (a \arr c) \dia (b \arr c)$.
\end{itemize}

\noindent
Adding distributivity is motivated by the desire to make the trees
\code{("a" $\,\hdia\,$ "b") $\,\harr\,$ "c"} and
\code{("a" $\,\harr\,$ "b") $\,\hdia\,$ ("a" $\,\harr\,$ "c")} mean the same
simplicial set (the rightmost one above).

Now that we know the rules of interpreting trees over the Boolean semiring, we
need to choose a target representation for simplicial sets. Alas, there is no
standard \code{Data.SimplicialS}\code{et}, so we need to come up with our own.
Here is a simple candidate:

@@

A key advantage of the \hcode{Tree}-based representation is that it is free from
any internal invariants, i.e. any value of \hcode{Tree Bool a} describes a
(valid) simplicial set. \hcode{SimplicialSet} does not satisfy this property;
for example, the set with two lists \code{[a]} and \code{[a,b]} is not a
simplicial set, because the edge $\textit{ab}$ appears in the set without its
vertex $\textit{b}$.

@@

\subsection{Other kinds of graphs}

>>>>

% Fix broken footnotes
\makeatletter
\renewcommand\@makefntext[1]{\leftskip=0.7em\hskip-0.75em\@makefnmark\hspace{1mm}#1}
\makeatother
\renewcommand{\footnotesize}{\small}
\setlength{\footnotesep}{3mm}

@@

\lstset{
  backgroundcolor=\color{white},     % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}; should come as last argument
  % identifierstyle=\color{red},
  basicstyle=\normalsize\ttfamily,      % the size of the fonts that are used for the code
  breakatwhitespace=false,              % sets if automatic breaks should only happen at whitespace
  breaklines=false,                     % sets automatic line breaking
  captionpos=b,                         % sets the caption-position to bottom
  abovecaptionskip=-3 mm,
  commentstyle=\itshape\color{graygreen}, % comment style
  % escapeinside={(:}{:)},             % if you want to add LaTeX within your code
  escapechar={!},
  % extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  % firstnumber=1000,                % start line enumeration with line 1000
  % frame=tb,                        % adds a frame around the code
  % keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{darkblue},     % keyword style
  language=Haskell,                  % the language of the code
  morekeywords={ Set, Tree, Leaf, Node, Applicative, fmap, liftA2, bimap, foldMap
               , traverse, mappend, pure, Foldable, Traversable, zero, one
               , Semiring, Semigroup, NonEmpty, sconcat, TSet,
               , SimplicialSet, TSimplicialSet, Graph, TGraph, LGraph
               , Map, IsString, fromString },
  deletekeywords={instance, data, where, class, filter, type, insert, delete, union, map},      % if you want to delete keywords from the given language
  emph={data, class, instance, where, type},
  emphstyle=\color{darkpink},
  numbers=none,                      % where to put the line-numbers; possible values are (none, left, right)
  % numbersep=5pt,                   % how far the line-numbers are from the code
  % numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  % rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  % showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  % showstringspaces=false,          % underline spaces within strings only
  % showtabs=false,                  % show tabs within strings adding particular underscores
  % stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{grayred},     % string literal style
  % tabsize=2,                       % sets default tabsize to 2 spaces
  % title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
  xleftmargin=10pt,
  aboveskip=8pt,
  belowskip=4pt
}

@@

\newcommand{\add}{\text{\raisebox{-0.1mm}{\large $\oplus$}}}
\newcommand{\mul}{\text{\raisebox{-0.1mm}{\large $\otimes$}}}
\newcommand{\zero}{\raisebox{-0.2mm}{\textcircled{\textsf{0}}}\xspace}
\newcommand{\one}{\raisebox{-0.2mm}{\textcircled{\textsf{\hspace{-0.2mm}1}}}\xspace}
\newcommand{\two}{\raisebox{-0.2mm}{\textcircled{2}}\xspace}
\newcommand{\dia}{\,\text{\raisebox{-0.3mm}{\Large $\diamond$}}\,}
\newcommand{\arr}{\,\text{$\rightarrow$}\,}
\newcommand{\ldia}{\text{\raisebox{-0.3mm}{\Large $\diamond$}}}

@@

\title{United Monoids}
\subtitle{Finding Simplicial Sets and Labelled Algebraic Graphs in Trees}
\author[~]{Andrey Mokhov}
\affiliation[~]{\hspace{-5.3mm}Jane Street, London, UK}
\affiliation[~]{\hspace{-5.3mm}Newcastle University, Newcastle upon Tyne, UK}

@@

\emph{Approach:} We follow a fairly standard approach of parameterising a data
structure with a semiring of edge labels. The newly developed data structure is
both more general and simpler: the two ways of composing graphs used in the
original formulation can now be obtained from a single operation by fixing the
semiring parameter to zero and one, respectively.

\emph{Knowledge:} By analysing the resulting data structure and formulating laws
for interpreting expressions of the corresponding type, we discover an unusual
algebraic structure, which we call ``united monoids'', that is, a pair of
monoids whose unit elements coincide. We believe that it is worth studying
united monoids in their full generality, going beyond the graphs which prompted
their discovery. To that end, we characterise united monoids with a minimal set
of axioms, prove a few basic theorems, and discuss several notable examples.

@@

Graphs and various graph-like combinatorial structures are ubiquitous in
programming. There are numerous approaches to representing graphs, from good old
adjacency matrices~\cite{cormen2009introduction} to categorical graph
algebras~\cite{2010_selinger_survey}. This paper introduces a new purely
functional graph representation, building on the earlier work on
\emph{algebraic graphs}~\cite{mokhov_alga}. The new representation is based on
elementary foundations (trees and semirings), and provides an expressive
language for modelling graph-like structures.

@@

\begin{lstlisting}
size :: Tree s a -> Int
size = fold (\a -> 1) (\s x y -> x + y)
\end{lstlisting}

\noindent
Here we ignore the labels of leaves and internal nodes: every leaf is
interpreted as~$1$, and every internal node simply adds up the sizes of its
children. In the rest of the paper, we will study more interesting ways to fold
trees. In particular, we will see how the very same data type can serve us when
working with sets, preorders, various flavours of graphs, and simplicial sets.
More specifically, our contributions are as follows.

@@

\vspace{-1mm}
\begin{itemize}
  \item The lack of connectivity is denoted by~\zero. For instance, zero network
        bandwidth.
  \item Addition~$\add$ is a commutative\footnote{
            An operation $\bullet$ is \emph{commutative} if
            $a \bullet b = b \bullet a$.
        } operation for combining connections in \emph{parallel}; e.g.,
        selecting the \emph{maximum} bandwidth link between two locations.
  \item Multiplication~$\mul$ is for combining connections in \emph{sequence}.
        Continuing the example, $\mul$ corresponds to taking the \emph{minimum}
        bandwidth of network links on route between two locations.
        Multiplication distributes\footnote{
            \emph{Distributivity} of $\otimes$ over $\oplus$ means
            $a \otimes (b \oplus c) = (a \otimes b) \oplus (a \otimes c)$ and
            $(a \oplus b) \otimes c = (a \otimes c) \oplus (b \otimes c)$.
        } over addition and has \zero as a zero\footnote{
            An element $z$ is a \emph{zero} (or an \emph{annihilating element})
            of an operation $\bullet$ if $z \bullet a = a \bullet z = z$.
        }.
  \item Finally, \one denotes either a unit of connectivity or infinite
        connectivity, depending on a specific semiring. In the bandwidth
        semiring, we have $\text{\one} = \infty$.
\end{itemize}

@@

class Semiring s where -- In this paper, !\h{s}! in !\h{Tree s a}! is always a !\h{Semiring}!
    zero :: s !\hspace{14.9mm}!-- Unit of !\hadd!, zero of !\hmul!
    (!\hadd!) !\hspace{1.55mm}!:: s -> s -> s !\hspace{1.4mm}!-- Associative, commutative
    one !\hspace{0.75mm}!:: s !\hspace{14.9mm}!-- Unit of !\hmul!
    (!\hmul!) !\hspace{1.55mm}!:: s -> s -> s !\hspace{1.4mm}!-- Associative, distributes over !\hadd!

@@

\begin{lstlisting}[float,label=lst-tree,xleftmargin=0pt,belowskip=-2mm,caption={
    A basic API for constructing and manipulating trees. We use the standard
    module \code{Data.L}\code{ist.N}\code{onEmpty} for representing non-empty
    lists, and \code{Data.S}\code{et} for sets.}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
leaf :: a -> Tree s a -- A tree with a single leaf
leaf = Leaf

leaves :: Semiring s => NonEmpty a -> Tree s a -- A tree of !\n{``}!disconnected!\n{''}! leaves
leaves = sconcat . NonEmpty.map Leaf -- Via !\h{Semigroup}!'s !\h{sconcat}!

node :: s -> Tree s a -> Tree s a -> Tree s a -- Combine two trees into an !\h{s}!-labelled node
node = Node

size :: Tree s a -> Int -- The number of leaves in a tree
size = length -- Via !\h{Foldable}!

leafSet :: Ord a => Tree s a -> Set a -- The set of leaves of a tree
leafSet = fold Set.singleton (const Set.union)

hasLeaf :: Eq a => a -> Tree s a -> Bool -- Test if a tree contains a leaf !\h{a}!
hasLeaf = elem -- Via !\h{Foldable}!

prune :: Tree s (Maybe a) -> Maybe (Tree s a) -- Prune all leaves labelled with !\h{Nothing}!
prune = sequence -- Via !\h{Traversable}!

filter :: (a -> Bool) -> Tree s a -> Maybe (Tree s a) -- Prune all leaves !\h{a}! with !\h{p a = False}!
filter p = traverse (\a -> if p a then Just a else Nothing) -- Via !\h{Traversable}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

\noindent
It is sometimes necessary to be able to express the \emph{empty tree}. To do
that, we simply wrap the \hcode{Tree} data type into \hcode{Maybe}. This is
illustrated by the function \hcode{prune} (Listing~\ref{lst-tree}), which prunes
all leaves labelled with \hcode{Nothing} and returns a possibly empty tree as a
result. By combining the empty tree and the operation \dia, we can extend our
type of trees from a semigroup to a \emph{monoid}, i.e., a semigroup with a unit
element.

When tree interpretations respect the associativity of \dia, \zero-labelled
trees become isomorphic to non-empty lists. This is evidenced by the functions
\hcode{toList} (provided by the \hcode{Foldable} instance in
Listing~\ref{lst-tree-std}) and \hcode{leaves} (Listing~\ref{lst-tree}). In the
subsequent sections, we will study data structures obtained by instantiating
\hcode{Tree s a} with different semirings~\hcode{s}, and by adding more
requirements (on top of associativity) to tree interpretations.

@@

\noindent
As hinted by the type synonym's name, we will interpret values \hcode{TSet a} as
(non-empty) sets; to get a standard \hcode{Set a} from a \hcode{TSet a}, we can
simply reuse \hcode{leafSet} from Listing~\ref{lst-tree}.

\begin{lstlisting}[label=lst-set,xleftmargin=0pt,belowskip=7pt,caption={
    Implementing a part of the standard \code{Data.S}\code{et} API with
    \hcode{TSet}.
}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
type TSet a = Tree () a -- Sets are trees with !\h{()}!-labelled internal nodes

instance Semiring () where -- The trivial semiring with !\h{zero = one = ()}!
    zero !\hspace{2.55mm}!= ()
    () !\add! () = ()
    one !\hspace{3.55mm}!= ()
    () !\mul! () = ()

singleton :: a -> TSet a !\hspace{32.18mm}\vrule\hspace{8mm}!insert :: a -> TSet a -> TSet a
singleton = Tree.leaf    !\hspace{35.1mm}\vrule\hspace{8mm}!insert = (!\hdia!) . singleton -- Via !\h{Semigroup}!

delete :: Eq a => a -> TSet a -> Maybe (TSet a) -- The empty set is represented by !\h{Nothing}!
delete a = Tree.filter (/= a)

member :: Eq a => a -> TSet a -> Bool !\hspace{9.1mm}\vrule\hspace{8mm}!union :: TSet a -> TSet a -> TSet a
member = Tree.hasLeaf                 !\hspace{30.33mm}\vrule\hspace{8mm}!union = (!\hdia!) -- Via !\h{Semigroup}!

size :: Ord a => TSet a -> Int
size = Set.size . leafSet -- We can't use !\h{Tree.size}! because it uses a non-idempotent !\h{fold}!

cartesianProduct :: TSet a -> TSet b -> TSet (a, b)
cartesianProduct = liftA2 (,) -- Via !\h{Applicative}!

filter :: (a -> Bool) -> TSet a -> Maybe (TSet a) -- The result can be empty
filter = Tree.filter

map :: (a -> b) -> TSet a -> TSet b
map = fmap -- Via !\h{Functor}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

\noindent
Listing~\ref{lst-set} shows how to implement a part of the standard set API by
reusing the more general \hcode{Tree}-manipulating functions defined earlier
in~\S\ref{sec-tree}.

In addition to associativity, the set interpretation of trees puts two new
requirements on every call site of \hcode{fold leaf node}\footnote{
    Haskell's type system isn't powerful enough to check such requirements, so
    we should use \hcode{fold} with care. A dependently typed language, such as
    Agda~\cite{2007_norell_agda}, could help us here.
}:

@@

\section{Simplicial sets}\label{sec-simplicial-set}
\vspace{-1mm}

@@

A \emph{simplicial set}~\cite{friedman_simplicial_sets} is a set of
\emph{simplices} of various dimensions along with their incidence relation.
A \emph{$0$-simplex} is a just a point, or a \emph{vertex}, and will correspond
to a leaf in our trees. A \emph{$1$-simplex} is a pair of connected vertices, or
an \emph{edge}; a \emph{$2$-simplex} is a filled-in \emph{triangle}; a
\emph{$3$-simplex} is a solid \emph{tetrahedron}; and so on. Simplicial sets
model systems that cannot be decomposed into pairwise relationships of their
components. For example, it can be useful to distinguish three truly
concurrent events from three events that are concurrent pairwise but constrained
by shared resources~\cite{rosenblum1989look}, e.g., three people eating
ice-cream with two spoons. The former corresponds to a $2$-simplex, while the
latter -- to a simplicial set with three $1$-simplices (a hollow triangle).

To represent simplices of dimension higher than $0$, we will \emph{connect}
their vertices using \one-labelled internal nodes. The figure below shows the
simplicial sets corresponding to the trees
\code{"a"}, \code{"a" $\,\harr\,$ "b"},
\code{("a" $\,\harr\,$ "b") $\,\harr\,$ "c"} and
\code{("a" $\,\hdia\,$ "b") $\,\harr\,$ "c"}, where the operator \code{$\harr$}
is defined as \hcode{x}~\code{$\harr$}~\hcode{y}~\code{=}~\hcode{node one x y}
(see Listing~\ref{lst-simplicial-set}).

\vspace{2.5mm}
\hfill\hspace{-4mm}\includesvg[scale=0.28]{simplicial-set-examples}\hfill
\vspace{2.5mm}

\noindent
Note the difference between the two last examples: by changing the label of one
of the nodes from \one to \zero, we break the corresponding connection,
and as a result, a $2$-simplex (triangle $\textit{abc}$) falls apart into
one-dimensional simplices (edges $\textit{ac}$~and~$\textit{bc}$).

@@

\vspace{-1mm}\begin{itemize}
    \item Associativity, inherited from~\S\ref{sec-tree}:

          \vspace{0.5mm}\hspace{10mm}$(a \dia b) \dia c = a \dia (b \dia c)$~~and~~$(a \arr b) \arr c = a \arr (b \arr c)$;

    \item Commutativity and idempotence of \dia, inherited from~\S\ref{sec-set}:

          \vspace{0.5mm}\hspace{10mm}$a \dia b = b \dia a$~~and~~$a \dia a = a$;

    \item Containment (\textbf{\color{darkblue}new!}), i.e., every simplex
          \emph{contains} its sub-simplices:

          \vspace{0.5mm}\hspace{10mm}$a \arr b = (a \arr b)\ \dia\ a\ \dia\ b$;

    \item Distributivity (\textbf{\color{darkblue}new!}):

          \vspace{0.5mm}\hspace{10mm}$a \arr (b \dia c) = (a \arr b) \dia (a \arr c)$~~and~~$(a \dia b) \arr c = (a \arr c) \dia (b \arr c)$.
\end{itemize}

\noindent
Adding distributivity is motivated by the desire to make the trees
\code{("a" $\,\hdia\,$ "b") $\,\harr\,$ "c"} and
\code{("a" $\,\harr\,$ "c") $\,\hdia\,$ ("b" $\,\harr\,$ "c")} mean the same
simplicial set (the rightmost one above).

\begin{lstlisting}[float,label=lst-simplicial-set,xleftmargin=0pt,caption={
    A basic API for working with simplicial sets represented by trees.
}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
type TSimplicialSet a = Tree Bool a -- Simplicial sets are trees over the Boolean semiring

instance Semiring Bool where -- The Boolean semiring with !\h{zero = False}! and !\h{one = True}!
    zero = False
    (!\hadd!) !\hspace{1.5mm}!= (||)
    one !\hspace{0.7mm}!= True
    (!\hmul!) !\hspace{1.5mm}!= (&&)

(!\harr!) :: TSimplicialSet a -> TSimplicialSet a -> TSimplicialSet a
(!\harr!) = Tree.node one

overlay :: TSimplicialSet a -> TSimplicialSet a -> TSimplicialSet a
overlay = (!\hdia!) -- Via !\h{Semigroup}!; recall that !\h{(\hdia) = Tree.node zero}!

connect :: TSimplicialSet a -> TSimplicialSet a -> TSimplicialSet a
connect = (!\harr!)

vertex :: a -> TSimplicialSet a -- 0-simplex
vertex = Tree.leaf

edge :: a -> a -> TSimplicialSet a -- 1-simplex
edge x y = connect (vertex x) (vertex y)

triangle :: a -> a -> a -> TSimplicialSet a !\hspace{6.2mm}!-- 2-simplex
triangle x y z = connect (vertex x) (edge y z) -- Equivalently, !\h{connect (edge x y) (vertex z)}!

simplex :: NonEmpty a -> TSimplicialSet a -- N-simplex
simplex = foldr1 connect . NonEmpty.map vertex -- Via !\h{Foldable}!

hasVertex :: Eq a => a -> TSimplicialSet a -> Bool
hasVertex = Tree.hasLeaf

vertexSet :: Ord a => TSimplicialSet a -> Set a
vertexSet = Tree.leafSet

filter :: (a -> Bool) -> TSimplicialSet a -> Maybe (TSimplicialSet a)
filter = Tree.filter

map :: (a -> b) -> TSimplicialSet a -> TSimplicialSet b -- The so-called !\n{``}!simplicial map!\n{''}!
map = fmap -- Via !\h{Functor}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

\noindent
Now that we know the rules of interpreting trees over the Boolean semiring, we
need to choose a target representation for simplicial sets. Alas, there is no
standard \code{Data.SimplicialS}\code{et}, so we need to come up with our own.
Here is a simple candidate:

@@

A key advantage of the \hcode{Tree}-based representation is that it is free from
any internal invariants, i.e., any value of \hcode{Tree Bool a} describes a
(valid) simplicial set. \hcode{SimplicialSet} does not satisfy this property;
for example, the set with two lists \code{[a]} and \code{[a,b]} is not a
simplicial set, because the edge $\textit{ab}$ appears in the set without its
vertex $\textit{b}$.

@@

\subsection{Other kinds of graphs}\label{sec-other-graphs}
